Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models

### 2024-08-30 manuscript.Rmd
Update manuscript

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of
biological and cultural traits over evolutionary time. Among these methods, 
some of the most popular are phylogenetic generalised linear mixed models, 
phylogenetic path analysis, and Pagel's 'discrete' method. While these methods
have clear benefits under specific scenarios, they also have important
limitations, such as unrealistic assumptions about the evolutionary process,
a focus on evolutionary correlation rather than causation, and limits on the 
number and type of traits that researchers can include. Here, we develop a
novel class of Bayesian *dynamic phylogenetic models* (DPMs) that overcomes
these limitations and synthesizes the strengths of existing phylogenetic 
comparative methods into a single generalizable framework for dynamic
phylogenetic causal inference. We formally and conceptually introduce this
class of model, which is applicable to any number of traits following any
response distribution. We provide a short tutorial for fitting these models
with the `coevolve` R package, a user-friendly package purpose-built for 
fitting DPMs in Stan. We highlight previous studies that have used DPMs to 
make inferences about trait coevolution, including analyses of brain size 
in non-human primates and social complexity in human societies, and suggest
some directions for future model development and research.
\end{center}

@@

\begin{equation}
  \begin{aligned}
    \text{spermsize}_t &= (0.90 \times \text{spermsize}_{t-1}) + 
    (0.85 \times \text{promiscuity}_{t-1}) + \epsilon_{\text{spermsize}} \\
    \text{promiscuity}_t &= (0.90 \times \text{promiscuity}_{t-1}) + 
    \epsilon_{\text{promiscuity}} \\
    \epsilon_{\text{spermsize}}, \epsilon_{\text{promiscuity}} &\sim 
    \mathcal{N}(0,0.05)
  \end{aligned}
\end{equation}

@@

To determine whether this pattern of results generalizes beyond a single 
simulation run, we iterated the above simulation 100 times for three different 
sample sizes: 20, 50, and 100 species. Table \@ref{tab:power} shows the 
proportion of models where the posterior $\Delta\theta_z$ values have 95% 
credible intervals excluding zero. The results show that the model is relatively 
underpowered to detect the causal effect of promiscuity on sperm size when there
are only 20 species, but it is able to detect this causal effect with sufficient
power when the sample size is 50 or 100.

>>>>

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of
biological and cultural traits over evolutionary time. Among these methods, 
some of the most popular are phylogenetic generalised linear mixed models, 
phylogenetic path analysis, and Pagel's `discrete' method. While these methods
have clear benefits under specific scenarios, they also have important
limitations, such as unrealistic assumptions about the evolutionary process,
a focus on evolutionary correlation rather than causation, and limits on the 
number and type of traits that researchers can include. Here, we develop a
novel class of Bayesian \textit{dynamic phylogenetic models} (DPMs) that
overcomes these limitations and synthesizes the strengths of existing 
phylogenetic comparative methods into a single generalizable framework for 
dynamic phylogenetic causal inference. We formally and conceptually introduce
this class of model, which is applicable to any number of traits following any
response distribution. We provide a short tutorial for fitting these models
with the \texttt{coevolve} R package, a user-friendly package purpose-built for 
fitting DPMs in Stan. We highlight previous studies that have used DPMs to 
make inferences about trait coevolution, including analyses of brain size 
in non-human primates and social complexity in human societies, and suggest
some directions for future model development and research.
\end{center}

@@

\begin{equation}
  \label{eq:generative}
  \begin{aligned}
    \text{spermsize}_t &= (0.90 \times \text{spermsize}_{t-1}) + 
    (0.85 \times \text{promiscuity}_{t-1}) + \epsilon_{\text{spermsize}} \\
    \text{promiscuity}_t &= (0.90 \times \text{promiscuity}_{t-1}) + 
    \epsilon_{\text{promiscuity}} \\
    \epsilon_{\text{spermsize}}, \epsilon_{\text{promiscuity}} &\sim 
    \mathcal{N}(0,0.05)
  \end{aligned}
\end{equation}

@@

To determine whether this pattern of results generalizes beyond a single 
simulation run, we iterated the above simulation 100 times for three different 
sample sizes: 20, 50, and 100 species. Table \@ref{tab:power} shows the 
proportion of models where the posterior $\Delta\theta_z$ values have 95% 
credible intervals excluding zero. The results show that the model is relatively 
underpowered to detect the causal effect of promiscuity on sperm size when there
are only 20 or 50 species, but it is able to detect this causal effect with 
sufficient power when the sample size is 100. Note, however, that these power
estimates apply only to the generative model in \autoref{eq:generative};
we would expect to find different power estimates when simulating with different
effect sizes and variables with different response distributions.
